---
title: "Covid-19 confirmed cases weekly report (applied epi skill test)"
author: "Berhe Tesfay"
date: '2022-04-07'
output:
  pdf_document: default
  html_document: default
  word_document: default
---
```{r setup,include=FALSE,  results='hide', message=FALSE, warning=FALSE}
## hide all code chunks in the output, but show errors
knitr::opts_chunk$set(echo = FALSE,       # hide all code chunks in output
                      error = TRUE,       # show errors if they appear, but don't stop
                      fig.width = 6*1.25, # Figure width
                      fig.height = 6,      # Figure height
                      warning = FALSE,
                      message = FALSE
                     )

## set the environment TZ
Sys.setenv(TZ='GMT')

## Installing required packages for this template
required_packages <- c("knitr", # create output docs
                       
                       "here",        # find your files
                       "janitor",
                       "flextable",
                       "dplyr",       # clean/shape data
                       "forcats",     # clean/shape data
                       "stringr",     # clean text
                       "rio",         # read in data
                       "ggplot2",     # create plots and charts
                       "patchwork",   # combine plots in one
                       "linelist",    # Functions for cleaning/standardising data/dates
                       "matchmaker",  # dictionary-based standardization of variables
                       "incidence",   # create epicurves
                       "aweek",       # define epi weeks
                       "epitrix",     # epi helpers and tricks
                       "sf",          # encode spatial vector data
                       "ggspatial",   # plot maps
                       "mondate",
                       "xts",         # moving \naverages
                       "zoo",         # moving \naverages
                       "classInt",    # specifying breaks for maps
                       "excel.link",  # opening password protected files
                       "askpass",     # opening password protected files
                       "tsibble",     # time series data
                       "slider",      # time series data
                       "tidyr",       # long/long adjustments to data
                       "gtsummary",  # make nice tables
                       "data.table",   # for taking last and first values from data frames
                       "patchwork",   # combining plots together
                       "TTR",        # calculate the moving average
                       "anytime",     # POSIX Date converter
                       "matrixStats", # standard deviation matrix calculator 
                       "tmaptools",   # for getting geocoordinates (lon/lat) based on place names
                       "ISOweek",
                       "growthrates",
                       "glue",
                       "ggplot2",
                       "scales",
                       "gridExtra",
                       "ggpubr",
                       "grid",
                       "sandwich",
                       "rgeos",
                       "countrycode",
                       "officer",
                       "gt",
                       "webshot",
                       "english",
                       "ggthemes",
                       "purrr",
                       "readxl",
                       "readr",
                       "broom",
                       "tidyverse",
                       "lmtest",
                       "parameters",
                       "see", 
                       "sitrep", # a package for outbreak analysis
                       "skimr") # skim the whole data set to start data cleaning abd after data cleaning
for (pkg in required_packages) {
  # install packages if not already present
  if (!pkg %in% rownames(installed.packages())) {
    install.packages(pkg)
  }
  
  # load packages to this current session 
  library(pkg, character.only = TRUE)
}

# set epi curve theme

epicurve_theme <- theme_classic() +
  theme(
    ## set the size for text 
    text = element_text(size = 12),
    legend.title=element_blank(),
    ## colour and size the grid lines in the plot 
    panel.grid.major = element_line(colour = "grey44",linetype = 3 ,size=1.0), 
    panel.grid.minor = element_line(colour = "grey44",linetype = 3, size =1.0), 
    ## add space between faceted plots (axis labels run in to each other)
    panel.spacing = unit(1, "lines"), 
    ## rotate x axis labels
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1,size=10),
   ## set the legend position
   legend.position = ("bottom"),
   legend.direction = ("horizontal")
  )

#devtools::install_github("crsh/citr") # insert citations thriugh addins

```


```{r fix the reprt period and the cut of weeks}
# Set the left and right censoring for date of consultation. 
# Remove the epi-weeks earlier than week 01/2020 as some of the dates are wrong
covid_start_week<- yearweek("2020 W01")

```


```{r import data}
setwd("C:/Users/berhe.tesfay/Desktop/Applied epi Application") # this is automatically created when you create your project 
# every thing is saved to the R-project folder with unique name and R gets it through the working directory
covid_fake_dat<-read_excel("covid_example_data.xlsx")

```

<!-------Skim the data fo start the data cleaning---------------->

```{r data exploration and cleaning}
# Skim the data to see all they
     #skim(covid_fake_dat)

# Also use the summary () almost the same out put but less structured 
      #summary(covid_fake_dat)
```

<!-------------clean the data------------------->

# Introduction

This analysis is a s sample analysis form a fake  covid-19 data for skill test.Neither the county mentioned here nor the data is realistic. 
there were a total of `r nrow(covid_fake_dat)` COVID-19 consultations in Fulton county of the united states between `r covid_start_week` and the date of reporting. Total number of confirmed cases reported were `r fmt_count(covid_fake_dat, covid_dx == "Confirmed")`.

```{r clean data for further analysis}
# check for duplicated cases(Cases PID seems the unique identifier)
   #duplicated(covid_fake_dat$case_PID) # this shows no duplicates (considering this as unique ID)

#1,this data doesn't have a dictionary so some of the variables  will be only assumed as they are not clear 
 ## clean the dates
 covid_fake_dat<-covid_fake_dat %>%
  mutate(date_onset=as.Date(sym_startdt_FALSE)) %>% 
  mutate(date_test_positive=as.Date(pos_sampledt_FALSE)) %>% 
  mutate(admission_date=as.Date(hosp_admidt_FALSE)) %>% 
  mutate(discharge_date=as.Date(hosp_dischdt_FALSE)) %>% 
  mutate(date_death=as.Date(died_dt_FALSE))
 
# in some cases we have the date of death but the date of admission is mission. so mutate the missing date of discharge as same as date of death.

# OR
       #covid_fake_dat<-covid_fake_dat %>% 
           #mutate(discharge_date = ifelse(is.na(discharge_date, date_death, discharge_date)))
# OR
       #covid_fake_dat<-covid_fake_dat %>% 
       #ifelse(is.na( covid_fake_dat$discharge_date), covid_fake_dat$date_death,  covid_fake_dat$discharge_date)
# OR
        #covid_fake_dat$discharge_date[is.na( covid_fake_dat$discharge_date)] <-  covid_fake_dat$date_death[is.na( #covid_fake_dat$discharge_date)]

# calculate the delay for health seeking behavior by calculating the delay for testing
       covid_fake_dat<-covid_fake_dat %>% 
          ## to include only the symptomatic cases (filter=="Symptomatic")
           dplyr::filter(Contact_id=="Yes-Symptomatic") %>% 
         ## calculate the date difference as number between data symptom start and date taste positive
           mutate(delay_in_detection=as.numeric(date_test_positive-date_onset))
         ## drop the negative values as the dealy cn not be retrospective and there is codding eror on the date

# calculate the length of admission for those hospitalized
       covid_fake_dat<-covid_fake_dat %>% 
         ## to include only the symptomatic cases (filter=="Symptomatic")
          dplyr::filter(hospitalized=="Yes") %>% 
          ## calculate the date difference as number between data symptom start and date taste positive
           mutate(length_admission=as.numeric(discharge_date-admission_date)) 
# calculate the length of days between admission and date of death
        covid_fake_dat<-covid_fake_dat %>% 
         ## to include only the symptomatic cases (filter=="Symptomatic")
          #dplyr::filter(died=="Yes" & died_covid=='Yes') %>% 
          ## calculate the date difference as number between data symptom start and date taste positive
           mutate(length_to_death=as.numeric(date_death-admission_date)) 
       
# Age of patient is not specified either in years or months (Assume this is in years)
# Then create age categories according to the WHO age category for COVID-19
## categorize  data Age 
covid_fake_dat<-covid_fake_dat %>% 
## age group for Continent wise analysis based on MSF Analysis plan
mutate(`Age group`=case_when(
# criteria                  # new value
case_age<2 ~"0-1 yrs",
case_age>=2 & case_age <5  ~"2-4 yrs",
case_age>=5 & case_age <15 ~"5-14 yrs",
case_age>=15 & case_age<30 ~"15-29 yrs",
case_age>=30 & case_age<40 ~"30-39 yrs",
case_age>=40 & case_age<50 ~"40-49 yrs",
case_age>=50 & case_age<60 ~"50-59 yrs",
case_age>=60 & case_age<70 ~"60-69 yrs",
case_age>=70 & case_age<80 ~"70-79 yrs",
case_age>=80 ~ "80+ yrs",
is.na(case_age) ~ NA_character_)) %>% 
#Mutate Gender
mutate(gender=case_gender) %>% 
mutate(race=case_race) %>% 
mutate(ethnicity=case_eth) %>% 

## Mutate the signs and symptoms
#create variable fever (based on the thermometer measurement)
mutate(measured_fever=sym_fever) %>% 
 mutate(measured_fever=factor(measured_fever,levels=list('Yes',
                                   'No',
                                    NA_character_))) %>% 
## Subjective fever (Based on the complain of the patient)
  mutate(subjective_fever=sym_subjfever) %>% 
 mutate(subjective_fever=factor(subjective_fever,levels=list('Yes',
                                   'No',
                                    NA_character_))) %>% 
## create variable cough
mutate(cough=sym_cough) %>% 
 mutate(cough=factor(cough,levels=list('Yes',
                                   'No',
                                    NA_character_))) %>% 
## create variable myalgia
mutate(myalgia=sym_myalgia) %>% 
 mutate(myalgia=factor(myalgia,levels=list('Yes',
                                   'No',
                                    NA_character_))) %>%   
  
## lose smell and taste
mutate(anosmia_agusia=sym_losstastesmell) %>% 
 mutate(anosimia_ageusia=factor(anosmia_agusia,levels=list('Yes',
                                   'No',
                                    NA_character_))) %>%  
## Sore throat
mutate(sorethroat=sym_sorethroat) %>% 
 mutate(sorethroat=factor(sorethroat,levels=list('Yes',
                                   'No',
                                    NA_character_))) %>%   
  
## headache
 mutate(headache=sym_headache) %>% 
 mutate(headache=factor(headache,levels=list('Yes',
                                   'No',
                                    NA_character_))) 
```

```{r prepare data for inline text}
male_to_female_Ratio<-covid_fake_dat %>% 
         mutate(male_female=(sum(covid_fake_dat$gender=="Male"))/(sum(covid_fake_dat$gender=="Female")))
Medain_age<-covid_fake_dat %>% dplyr::select(case_age) %>% tbl_summary(statistic = all_continuous()~c("{median}, {mean},IQR ({p25}-{p75}"))


```

# Socio demographics 
The number of female and male patients were `r fmt_count(covid_fake_dat, gender == "Male")` and `r fmt_count(covid_fake_dat, gender == "Female")`. The medain, mean and IQR of age of patients was `r inline_text(Medain_age, variable='case_age')` years. 


```{r crate summary variables for plot}
## Create epidemiological week of date of onset 
   covid_fake_dat<-covid_fake_dat %>% 
     mutate (epiweek=yearweek(date_onset)) %>% 
# filter epiweek to 2020 and later to drop the wrong dates from the 1950s and 80s
     filter(epiweek>=covid_start_week)

## create long data set 
weekly_cases<-covid_fake_dat %>% 
  select(epiweek, race, ethnicity, died,died_covid) %>% 
  group_by(epiweek,race,ethnicity,died,died_covid) %>% 
  tally() 
  
## create moving average of the confirmed cases  
weekly_2ma<-covid_fake_dat %>% 
  group_by(epiweek) %>% 
  tally() %>%  
  mutate(two_week_ma=round(slide_dbl(n, ~mean(.x, na.rm = TRUE),
                                          .before = 3 ), 
                                digits = 0))

##create moving average grouped by gender,race, ethnicity
weekly_2ma_gender<-covid_fake_dat %>% 
  group_by(epiweek,gender,race,ethnicity) %>% 
  tally() %>%  
  mutate(two_week_ma=round(slide_dbl(n, ~mean(.x, na.rm = TRUE),
                                          .before = 3 ), 
                                digits = 0))

```

## Epicurve of case per epidemiological week 

```{r age sex pyramid, fig.height=7, fig.width=14,fig.cap='Age sex pyramid of confirmed covid-19 patienys, from fake data'}

## plot the age sex pyramid
# first change age group in to factor
covid_fake_dat$`Age group`<-as.factor(covid_fake_dat$`Age group`)
# then plot the age sex pyramid
 age_sex_pyramid<-plot_age_pyramid(covid_fake_dat,
                                    age_group ="Age group", 
                                    split_by = "gender",
                                    na.rm = TRUE)+
                                    # exclude missing values
  labs(y = "Cases (n)", x = "Age group") + # change axis  labels
  theme(legend.position = "right",     # move legend to bottom
        legend.title = element_blank(), # remove title
        text = element_text(size = 14),  # change text size
        plot.title = element_text(size = 16)) +
  ggtitle(paste0("Age sex pyramid of cases from fake covid data"))

 
 ## alternatively
 missing_age_sex <- covid_fake_dat %>% summarise(age = sum(is.na(case_age)), sex = sum(is.na(gender)))

tbl_pyramid_confirmed_Continent <- covid_fake_dat %>% 
  drop_na(gender, case_age)%>% 
  dplyr::count( gender, `Age group`) %>% 
  mutate(n = if_else(gender == "Male", -n, n))

figure1 <- ggplot(tbl_pyramid_confirmed_Continent, aes(x = `Age group`, y = n, fill = gender)) +
  #facet_wrap(~Continent, ncol = 2, scales = "free_x") +
  geom_col() +
  geom_hline(yintercept = 0, colour = "black") +
  geom_text(aes(label = abs(n), hjust = if_else(n >= 0, 1.1, -0.1)), colour = "white", size = 4) +
  coord_flip() +
  scale_fill_manual(name = NULL, values = c("#FF007F", "blue","red1"), breaks = c("Male", "Female",""), labels = c("Males", "Females","")) +
  #scale_y_continuous(label = abs, limits = pyramid_limits) + 
  theme(legend.position = "bottom") +
  labs(x = "Age group",size=10, y = "Confirmed and admitted Patients", caption = glue::glue("Missing Data: Age {missing_age_sex$age}, Sex {missing_age_sex$sex}"))+ 
  theme(axis.text = element_text(size = 10))+
  theme(axis.title = element_text(size = 10))+
ggtitle(paste0("Age sex pyramid of cases from fake covid data"))

```



```{r plot fill by gender, fig.width=14, fig.height=7,fig.cap="Number of confirmed cases and two weeks moving average"}

epi_curve<-ggplot() +
  geom_col(data = weekly_cases, 
           aes(x = epiweek, 
               y = n, 
               fill = died)) +
## add two week moving \naverage
  geom_line(data = weekly_2ma,
            aes(x = epiweek, 
                y = two_week_ma, colour = " 2 moving \naverage of ll cases"),
            size = 1.4)+
  scale_colour_manual( values = c("2 moving \naverage of ll cases" = "gray"))+

## rotate x axis text
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "bottom", # move legend to bottom
        #legend.title = element_blank(), # remove title
        text = element_text(size = 14),
        plot.title = element_text(size = 16))+  # change text size
  labs(title = "weely number of cases and biweely movng average",
       x = "epiweek",
       y = " Total number of cases",
       fill = " All deaths",
       colour = "")+
   scale_x_yearweek(date_breaks = "3 month", date_minor_breaks = "2 week",
             date_labels = "W%W")+
theme(legend.direction = "vertical",
legend.box = "horizontal")


## Died because of covid
epi_curve_died_covid<-ggplot() +
  geom_col(data = weekly_cases, 
           aes(x = epiweek, 
               y = n, 
               fill = died_covid)) +
## add two week moving \naverage
  geom_line(data = weekly_2ma,
            aes(x = epiweek, 
                y = two_week_ma, colour = " 2 moving \naverage of ll cases"),
            size = 1.4)+
  scale_colour_manual( values = c("2 moving \naverage of ll cases" = "gray"))+

## rotate x axis text
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "bottom", # move legend to bottom
        #legend.title = element_blank(), # remove title
        text = element_text(size = 14),
        plot.title = element_text(size = 16))+  # change text size
  labs(title = "weely number of cases and biweely movng average",
       x = "epiweek",
       y = " Total number of cases",
       fill = "Covid Death",
       colour = "")+
   scale_x_yearweek(date_breaks = "3 month", date_minor_breaks = "2 week",
             date_labels = "%d %b %Y\n(W%W)")+
theme(legend.direction = "vertical",
legend.box = "horizontal")




## combine the two plots 
epi_curve  +epi_curve_died_covid + figure1
 
```




# Signs and symptoms
The commonest signs and symptoms among the confirmed cases was cough `r fmt_count(covid_fake_dat, cough == "Yes")` followed by fever `r fmt_count(covid_fake_dat, measured_fever == "Yes")` and Myalgia `r fmt_count(covid_fake_dat, myalgia == "Yes")`(Table 1). 


```{r summary of signs and symptoms}
ssymptoms<-covid_fake_dat %>% 
  select(subjective_fever, measured_fever,cough,sorethroat,anosmia_agusia,myalgia,headache, gender) %>% 
  tbl_summary(by=gender,
              missing='no',
             statistic=all_categorical()~c("{n}/{N}({p}%)"),
             value = list(subjective_fever~"Yes",
                          measured_fever~"Yes",
                          cough~"Yes",
                          sorethroat~"Yes",
                          anosmia_agusia~"Yes",
                          myalgia~"Yes",
                          headache~"Yes"),
             label = list(subjective_fever~"Subjective fever",
                          measured_fever~"Measured fever",
                          cough~"Cough",
                          sorethroat~"Sore throat",
                          anosmia_agusia~"Anosmia/Ageusia",
                          myalgia~"Myalgia",
                          headache~"Headaceh")) %>% 
  add_n() %>% 
  add_overall(last=T) %>% 
  modify_caption("Signs and symptoms of patienst, from fake covid data")
   theme_gtsummary_compact()
   theme_gtsummary_journal()
   theme_gtsummary_printer(print_engine = "flextable")
ssymptoms 
``` 
# Cases by ethnicity and race

Majority race of the patients were black and Non-Hispanic/Latinos take (figure 2). 

```{r plot cases by race and ethinicity, fig.height 7}
x<-ggplot(data = weekly_cases,
       aes(x=epiweek,
           y=n)) +
  geom_col(fill='blue')+
  geom_line(data = weekly_2ma_gender,
            aes(x=epiweek,
                y=two_week_ma,colour = "moving \naverage"),
            size = 1.2)+
  facet_wrap(race ~ ethnicity,nrow=4, scales = "free_x")+ 
  #facet_grid(race ~ ethnicity, scales = "free_x")+ 
  labs(title = "Number of confirmed COVID-19 patients by epidemiological week",
       x = "Week",
       y = "Number of confirmed cases and 2 moving avrage of all cases",
       fill = "Gender")+
  epicurve_theme
```


```{r fill the plots by race, ethnicty and covid mortality}
ggplot()+
  geom_col(data = weekly_cases, 
           aes(x = epiweek, 
               y = n, 
               fill = died_covid)) +
## add four week moving \naverage
  facet_grid(race ~ ethnicity, scales = "free_x")+
  scale_colour_manual( values = c("2 moving \naverage" = "green1"))+
  labs(title = "Number of confirmed COVID-19 patients by race and ethnicity",
       x = "Week",
       y = "Total confirmed cases for each strata",
       fill = "Covid death")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1,size=8),
        strip.text.y = element_text(size = 6, colour = "red1", angle = 90),
        strip.text.x = element_text(size = 8, colour = "blue"))
  
```



# Location of cases(heatmaps)


```{r plot heat map}

library(tidyverse)
library(stringr)
library(sf)
library(leaflet)
library(maps)
# the sf is saved to the project folder and can easily be imported using the read_sf package
map_fulton_county_raw <- read_sf("FultonCountyZipCodes.shp") %>% 
# clean anmes 
janitor::clean_names()


## check the coordinate reference system (CRS)
   # st_crs(map_fulton_county_raw)

## if CRS not WGS84, reset it WGS84
     map_fulton_county_raw <- st_set_crs(map_fulton_county_raw, value = 4326) # Sets to WGS84

# change the sf to data frame
    map_fulton_county_raw<-map_fulton_county_raw %>% 
# create variable for joinging
    mutate(case_zip=zip_code)

# create summary data of the covid case 
    covid_fake_dat$case_zip<-as.character(covid_fake_dat$case_zip)
    
   covid_fake_dat_confirmed<-covid_fake_dat %>% 
     filter(covid_dx=="Confirmed") %>% 
     group_by(case_zip) %>% 
      tally()
# mutate the merging variable in to character
  

# join the two data sets
    map_joined<-left_join(covid_fake_dat_confirmed,map_fulton_county_raw, by='case_zip')

map_joined<-st_as_sf(map_joined)

mab_joined<-st_set_crs(map_joined, value = 4326) # Sets to WGS84

```
 

```{r plot the heat map}

map<- ggplot() +    ## add in the back ground tiles
  geom_sf(data = map_joined, aes(fill = (n))) +  # needed to avoid gridlines being drawn
  coord_sf(datum = NA) +  # add a scalebar
  annotation_scale()
map+
scale_fill_continuous(name="All Confirmed cases", 
            low = "lightgreen", high = "darkgreen",
            limits = c(0,250), 
            breaks=c(0,50,100,150,200), na.value = "grey50") +

          labs(title="Total number of confirmed cases of COVID-19 in \nFilton county united sates bewtween date x and y, from fake covid data")

```

